### 2018-01-04 submission/manuscript.Rmd
re-write of discussion

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of either of these two. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are strongly associated with carcinoma but may not be associated with adenoma. Although CRC-associated genera are associated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC. Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity, are highly supportive of this model. However, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5]. Conversely, using the present data, it is less likely that adenoma development fits this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community. The poor performance of the Random Forest models for classifying adenoma, based only on the microbiota, would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-associated genera. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

The associations between the microbiota and adenoma are less clear in part because many studies may not be powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation. In contrast to adenoma, our observations suggest that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].

>>>>

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These mouth-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.

The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation. A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair. Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance. However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5]. This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma. Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development. The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself. The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that tsmall changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of *Propionibacterium*, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination. The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.

The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis. Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to poly formation. In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].

### 2018-01-04 submission/manuscript.tex
re-write of discussion

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the RR of either of these two. Even though CRC-associated genera
increase the RR of carcinoma, they do not consistently increase the
relative risk of adenoma. This information suggests that these specific
genera are strongly associated with carcinoma but may not be associated
with adenoma. Although CRC-associated genera are associated with
carcinomas, our data show that by using the whole community,
classification models perform better than when only the CRC-associated
genera are included. CRC-associated genera are clearly important to
carcinoma pathogenesis but accounting for the community in which these
microbes exist can drastically increase the ability of models to make
predictions.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased
carcinoma RR of CRC-associated genera versus \(\alpha\)-diversity
metrics and increasing RR with more CRC-associated genera positivity,
are highly supportive of this model. However, the initial establishment
of the driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The step-wise decrease in diversity
suggests that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma, based only on the microbiota, would suggest
that this is the case. One potential hypothesis from these observations
is that at early stages of the disease, how the host interacts with
these subtle changes is what ultimately leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
carcinoma. Based on these observations, it is possible to hypothesize
that the early changes in the community may give rise to initial polyp
formation via interactions with the host and not necessarily via
interactions with CRC-associated genera. These changes then create new
niches in which any one of the CRC-associated genera could gain a
foothold, exacerbating the initial changes in community and facilitating
the transition from adenoma to carcinoma via a driver-passenger type
mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with disease severity. For the OTU models
both the unmatched and matched {[}Figure S7E \& F{]} tissue samples had
some concordance with the stool data, with resident bacteria being the
most prevalent in the top 10 important variables across studies. Unlike
in stool, \emph{Fusobacterium} was the only CRC-associated bacteria
consistently present in the top 10 of the CRC models {[}Figure S7B-C \&
E-F{]}. The majority of the results seem to be study specific with many
top 10 taxa being present only in a single study. One other potentially
worrying sign is the presence of \emph{Propionibacterium} within the top
10 for the genera and OTU models and could be a marker of contamination.
The low biomass of these samples coupled with potential contamination
might be a possible reason why the tissue results seem to be more
sporadic then the stool results.

The associations between the microbiota and adenoma are less clear in
part because many studies may not be powered effectively to observe the
small changes reported here. None of the studies analyzed were properly
powered to detect a 10\% or lower change between case and controls. This
small effect size range may well be the scope in which differences
consistently occur in adenoma due to the subtle changes in community
that occur between control and adenoma. Future studies investigating
adenoma and the microbiota need to take these factors into consideration
if we are to work out how the microbiota contributes to adenoma
formation. In contrast to adenoma, our observations suggest that many
changes in carcinoma could easily result in effect sizes that are 30\%
or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 7{]}.

>>>>

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but ourobservations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of mouth-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these mouth-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of mouth-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
mouth-associated genera positivity {[}Figure 3C{]}, are highly
supportive of their importance. However, our observations show that when
adding the community context to our Random Forest models along with the
mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
This suggests the resident community cannot be ignored completely and
needs to be accounted for when investigating mouth-associated microbes
and carcinoma. Conversely, using the present data, it is less likely
that mouth-associated microbes are correlated with adenoma development.
The step-wise decrease in diversity suggests that the adenoma community
is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be disciminative across the
different studies for both adenoma and carcinoma {[}Figure 6{]}. This
data would suggest that whether the non-resident bacterium is
\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
how these bacteria interact with the changing resident community. Based
on these observations, it is possible to hypothesize that tsmall changes
in community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the OTU models of
unmatched and matched {[}Figure S7E \& F{]} tissue samples were
concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the tissue results seem to be study specific with many top
10 taxa being present only in a single study. Additionally, the presence
of \emph{Propionibacterium}, within the top 10 most important variables
for the genera and OTU models, is worrying because it is commonly a
marker of contamination. The low bacterial biomass of tissue samples
coupled with potential contaminationcould ultimately explain why the
these results seem to be more sporadic than the stool results.

The associations between the microbiota and adenoma are inconclusive, in
part, because many studies may not be powered effectively to observe
small effect sizes. None of the studies analyzed were properly powered
to detect a 10\% or lower change between cases and controls. A small
effect size may well be the scope in which differences consistently
occur in adenoma based on the results observed within our meta-analysis.
Future studies investigating adenoma and the microbiota need to take
power into consideration if we are to reproducibly study whether the
microbiota contributes to poly formation. In contrast to adenoma, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in carcinoma because of large effect size
differences between cases and controls {[}Figure 7{]}.
